---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Unzip data to obtain a csv file.

```{r unzip, cache=TRUE}
unzip("activity.zip",exdir = "data")
```
Reading the data into activity data frame and show some summary statistics
```{r reading, cache=TRUE}
activity <- read.csv("data/activity.csv", stringsAsFactors=FALSE)
str(activity)
summary(activity)
```

Convert date to POSIXct class using lubridate package and convert interval to hour:minute format
```{r date_conversion, message=FALSE}

activity$date <- as.Date(activity$date, format = "%Y-%m-%d")

```


## What is the average daily activity pattern?
#### 1. Calculate the total number of steps taken per day (ignore the missing values)
```{r daily_total,message=FALSE}
colMeans(is.na(activity))
steps_day <- aggregate(steps ~ date, rm.na = TRUE, data = activity, FUN = sum)
```

####  2. Make a histogram of the total number of steps taken each day

```{r histogram}
plot(steps_day, type = "h", lwd = 10, lend = "square")
```

#### 3. Calculate and report the mean and median of the total number of steps taken per day

```{r summary}
aggregate(steps ~ date, data = activity, FUN = mean)
aggregate(steps ~ date, data = activity, FUN = median)
```

## What is the average daily activity pattern?

#### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  

```{r daily}
plot(aggregate(steps ~ interval, data = activity, FUN = mean), type = "l")
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r daily2}
max(activity$steps, na.rm = TRUE)
```


## Imputing missing values

#### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r missing}
sum(is.na(activity))
```

#### 2. Devise a strategy for filling in all of the missing values in the dataset

As the number of missing values in this dataset is fairly large, we cannot be sure if there is no bias introduced by missing values. Therefore we impute missing values based on average number of steps in particular 5-minutes interval. 

#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r na_imputing, cache=TRUE}
activity2 <- activity
sapply(activity2, class)

activity2$steps[is.na(activity2$steps)] <- mean(na.omit(activity$steps))
activity2$date <- as.Date(activity2$date, format = "%Y-%m-%d")
```


#### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day

```{r histogram_no_NAs}
steps_day2 <- aggregate(steps ~ date, rm.na = TRUE, data = activity2, FUN = sum)

par(mfrow = c(1, 2))
plot(steps_day, type = "h", lwd = 5,lend = "square", main = "With NAs")
abline(h = seq(0, 20000, 2500), lty = "dashed")
plot(steps_day2, type = "h", lwd = 5, lend = "square", main = "NAs filled")
abline(h = seq(0, 20000, 2500), lty = "dashed")
dev.off()
```

##Filling the NA makes the distribution more homogeneous. By the way, this operation could hide interesting patterns such as the inactivity during particular days of the week.

``` {r}
aggregate(steps ~ date, data = activity, FUN = mean)
aggregate(steps ~ date, data = activity, FUN = median)
aggregate(steps ~ date, data = activity2, FUN = mean)
aggregate(steps ~ date, data = activity2, FUN = median)
```
## Are there differences in activity patterns between weekdays and weekends?

#### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day

```{r weekday}
activity2$weekday <- factor(format(activity2$date, "%A"))

levels(activity2$weekday) <- list(weekday = c("Monday", "Tuesday",
                                              "Wednesday", "Thursday",
                                              "Friday"), weekend =
                                          c("Saturday", "Sunday"))
```

#### 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)


```{r weekend_comparison}
par(mfrow = c(2, 1))

with(activity2[activity2$weekday == "weekend",], plot(aggregate(steps ~ interval, FUN = mean), type = "l", main = "Weekends"))

with(activity2[activity2$weekday == "weekday",], plot(aggregate(steps ~ interval, FUN = mean), type = "l", main = "Weekdays"))
dev.off()
```